---
title: "Earth Overshoot data 2024"
author: "Dr J P McKeown"
output: "html_document"
---

## Data fields for dashboard display: 
Country name,
Sustainability Grade (revised by EO),
Comment (revised by EO).
Sustainable maximum population (calculated as biocapacity/footprint),
Population (2023 from World Bank SP.POP.TOTL),
Population growth rate (most recent available from SP.POP.GROW),
Contraception (most recent Modern methods contraceptive prevalence % SP.DYN.CONM.ZS),
Species threatened (most recent IUCN quarterly),
GDP per person (most recent from World Bank NY.GDP.PCAP.CD),

## Data needed for calculation or processing:
iso3c
Biocapacity per person (GFN 2020 or estimated 2022)
Ecological Footprint per person (GFN 2020 or estimated 2022)

## Other national data fields available:
iso2c
Biocapacity (GFN)
Ecological Footprint (GFN)
Population (2023 from UN WPP)

```{r, setup}
library(tidyverse)
library(countrycode)
library(devtools)
options(timeout = 600)
install_github("PPgp/wpp2022")
library(WDI)
```

```{r, stem}
# make stem frame of Countries
eo_temp <- readRDS('data/eop_26April2022.rds')
eo_stem <- eo_temp %>%
  rename(Photos = Freq) %>% 
  relocate(Country, Continent) %>%
  relocate(Photos, .after = Continent) %>% 
  select(Country:iso3c) %>% 
  arrange(Country) # no effect as rds already sorted by country name

# official names of countries
eo_stem$Country[eo_stem$Country == "Democratic Republic of The Congo"] = "Democratic Republic of the Congo"
eo_stem$Country[eo_stem$Country == "Czech Republic / Czechia"] = "Czech Republic"
eo_stem$Country[eo_stem$Country == "Bosnia & Herzegovina"] = "Bosnia and Herzegovina"
eo_stem$Country[eo_stem$Country == "Trinidad & Tobago"] = "Trinidad and Tobago"
eo_stem$Country[eo_stem$Country == "Antigua & Barbuda"] = "Antigua and Barbuda"
eo_stem$Country[eo_stem$Country == "St. Kitts and Nevis"] = "Saint Kitts and Nevis"
eo_stem$Country[eo_stem$Country == "St. Lucia"] = "Saint Lucia"
eo_stem$Country[eo_stem$Country == "St Vincent and the Grenadines"] = "Saint Vincent and the Grenadines"
```

Merge data from Global Footprint Network: biocapacity, footprint, and population.
```{r, add GFN}
# GFN_2019 <- read_csv("data/GFN_2019.csv")
eo_2023_gfn <- eo_stem %>% 
  left_join(GFN_2019, by = "iso2c") %>% 
  rename(Biocapacity_pp_2019 = biocapacity_pp) %>% 
  rename(Footprint_pp_2019 = footprint_pp) %>%
  rename(Population_gfn_2019 = population) %>% 
  select(Country:iso3c, Biocapacity_pp_2019, Footprint_pp_2019, Population_gfn_2019)
# check France total population
eo_2023_gfn[eo_2023_gfn$iso3c == "FRA",] # 64,436,800
```

Calculate maximum sustainable population
```{r, sustainable}
eo_2023 <- eo_2023_gfn %>% 
  mutate(Sustainable_pop = floor(Biocapacity_pp_2019 / Footprint_pp_2019 * Population_gfn_2019))

```

Interim export for "Country Data" googlesheet
```{r}
# Country Comments are now in a permanent separate tab on googlesheet
write_csv(eo_2023, "data/eo_gfn_maxpop_July_2023.csv")
```

## IUCN threatened species total by country
Most recent is 2022-2 https://www.iucnredlist.org/support/whatsnew
```{r}
species_2022 = read_csv("data/IUCN_July2023download_Table_5 _Threatened_species_by_country.csv")
threat <- species_2022 %>%
  rename(Country = Name) %>% 
  rename(Species = Total) %>% 
  select(Country, Species)
head(threat)
```

Add iso3c based on name Country
```{r}
threat$iso3c <- countrycode(sourcevar = threat$Country, 
                        origin = "country.name",
                        destination = "iso3c")
head(threat)
sum(is.na(threat$iso3c)) # =1 "Disputed territory"
```

## United Nations data

https://github.com/PPgp/wpp2022
WPP2022 (Population) includes Taiwan

```{r}
library(wpp2022)
data(pop1dt)

pop21 <- pop1dt %>%
  rename(Population = pop) %>% 
  filter(year == 2021) %>%
  mutate(Population = Population * 1000) %>% 
  select(country_code, name, Population)

pop21$iso3c = countrycode(sourcevar = pop21$country_code,
                          origin = "iso3n",
                          destination = "iso3c")
UN_population_2021 <- pop21 %>% 
  filter(!is.na(iso3c))
# on inspection, only regions are Null iso3c (and Kosovo)

sum(is.na(UN_population_2021$Population)) # 0, all countries have 2021 data

UN_population_2021[UN_population_2021$iso3c == "FRA",] # 64,560,540
head(UN_population_2021)
```

Fertility rate (TFR) from UN WPP 2022
```{r}
data(tfr1dt)

TFR_2021 <- tfr1dt %>%
  filter(year == 2021) %>%
  select(country_code, tfr, name)

# maybe remove regions before running countrycode?

TFR_2021$iso3c = countrycode(sourcevar = TFR_2021$country_code,
                          origin = "iso3n",
                          destination = "iso3c")

# Inspection shows all countries have 2021 data; filter/slice not needed
```

Population Growth Rate % from UN WPP2022
```{r}
data(misc1dt)
grow21 <- misc1dt %>%
  rename(GrowthRate = growthrate) %>% 
  filter(year == 2021) %>% 
  filter(country_code < 900) %>%  # remove continental regions
  select(country_code, GrowthRate, name)

grow21$iso3c = countrycode(sourcevar = grow21$country_code,
                          origin = "iso3n",
                          destination = "iso3c")

sum(is.na(grow21$GrowthRate)) # 0, all countries have 2021 data
sum(is.na(grow21$iso3c)) # 1 missing
grow21[grow21$country_code == 412] # Kosovo not in iso

UN_PopGrowth_2021 <- grow21 %>% 
  filter(!is.na(iso3c)) %>%
  rename(PopGrowRate_UN_2021 = GrowthRate) %>% 
  select(iso3c, PopGrowRate_UN_2021)  

head(UN_PopGrowth_2021)
```

## World Bank data

Population total (compare with UN WPP)
```{r}
WB_pop_2021 <- WDI(
  country = 'all',
  indicator = 'SP.POP.TOTL',
  start = 2021,
  end = 2021,
  extra = FALSE,
  cache = NULL,
  latest = NULL,
  language = 'en'
)

# count missing rows = 0
sum(is.na(WB_pop_2021$SP.POP.TOTL))
# if>0, identify missing countries # =1 "Unclassified"

WB_pop_2021[WB_pop_2021$iso3c == "FRA",] # 67,749,632
head(WB_pop_2021)
```

```{r}
WB_pop_2022 <- WDI(
  country = 'all',
  indicator = 'SP.POP.TOTL',
  start = 2022,
  end = 2022,
  extra = FALSE,
  cache = NULL,
  latest = NULL,
  language = 'en'
)

# count missing rows = 0
sum(is.na(WB_pop_2022$SP.POP.TOTL))
# if>0, identify missing countries # =1 "Unclassified"

WB_pop_2022[WB_pop_2022$iso3c == "FRA",] # 67,935,660
head(WB_pop_2022)
```

Population Growth Rate, most recent
```{r, popgrowth}
# recent year and see if complete
popgrow_2022 <- WDI(
  country = 'all',
  indicator = 'SP.POP.GROW',
  start = 2022,
  end = 2022,
  extra = FALSE,
  cache = NULL,
  latest = NULL,
  language = 'en'
)

# count missing rows = 0
sum(is.na(pop_2021$SP.POP.GROW))
# if>0, identify missing countries

head(popgrow_2022)
```

Contraception, prevalence % modern methods
```{r, contraception}
tmp_contracept <- WDI(
  country = 'all',
  indicator = 'SP.DYN.CONM.ZS',
  start = 1984,
  end = 2022,
  extra = FALSE,
  cache = NULL,
  latest = NULL,
  language = 'en'
)
sum(is.na(tmp_contracept$iso3c)) # =0

contraception <- tmp_contracept %>%
  select(iso3c, year, SP.DYN.CONM.ZS, country) %>% 
  group_by(iso3c) %>% 
  filter(!is.na(SP.DYN.CONM.ZS)) %>% 
  slice_max(year)
  
head(contraception)
```

GDP per capita
```{r, GDP}
tmp_gdp <- WDI(
  country = 'all',
  indicator = 'NY.GDP.PCAP.CD',
  start = 2011,
  end = 2022,
  extra = FALSE,
  cache = NULL,
  latest = NULL,
  language = 'en'
)
sum(is.na(tmp_gdp$iso3c)) # =0

gdp_pp_2022 <- tmp_gdp %>%
  rename(Year_gdp = year) %>% 
  select(iso3c, Year_gdp, NY.GDP.PCAP.CD, country) %>% 
  group_by(iso3c) %>% 
  filter(!is.na(NY.GDP.PCAP.CD)) %>% 
  slice_max(Year_gdp)
  
head(gdp_pp_2022)
```

## Merge new data into Countries

### Dataframes to be merged into eo_2023:
popgrow_2022
UN_population_2021
contraception
threat(speacies)
gdp_pp_2022
TFR_2021

Merge pop growth rate % into EO country list data
```{r}
# eo_GFN_tmp <- eo_2023
popgrow_tmp <- popgrow_2022 %>% 
  select(-iso2c)

eo_2023 <- eo_GFN_tmp %>% 
  left_join(popgrow_tmp, by = "iso3c") %>% 
  rename(Growth_pop_2022 = SP.POP.GROW) %>% 
  select(Country:Sustainable_pop, Growth_pop_2022)
```

Merge UN population 2021
```{r}
eo_tmp <- eo_2023
eo_2023 <- eo_tmp %>% 
  left_join(UN_population_2021, by = "iso3c") %>%
  rename(Population_UN_2021 = Population)
  select(Country:Growth_pop_2022, Population_UN_2021)
eo_2023 <- eo_2023 %>% 
  select(-c(country_code, name))
head(eo_2023)
```

# could merge all columns in one codeblock, if incoming prepared beforehand
Contraception
```{r}
# rename contraception columns first
contracept_mergable <- contraception %>% 
  rename(Year_contracept = year) %>% 
  rename(Contraception = SP.DYN.CONM.ZS) %>% 
  select(iso3c, Contraception, Year_contracept)
```

Species threatened
```{r}
species_mergable <- threat %>%
  filter(!is.na(iso3c)) %>% 
  rename(Species_2022_2 = Species) %>% 
  select(iso3c, Species_2022_2)
```

GDP per capita
```{r}
gdp_mergable <- gdp_pp_2022 %>% 
  rename(GDP = NY.GDP.PCAP.CD) %>% 
  select(iso3c, GDP, Year_gdp)
```

TFR
```{r}
tfr_mergable <- TFR_2021 %>% 
  rename(TFR = tfr) %>% 
  select(iso3c, TFR)
```

Merge remaining columns
```{r}
eo_tmp_before_merge5 <- eo_2023
eo_2023 <- eo_tmp_before_merge5 %>% 
  left_join(UN_PopGrowth_2021, by = "iso3c") %>%
  left_join(contracept_mergable, by = "iso3c") %>%
  left_join(species_mergable, by = "iso3c") %>%  
  left_join(gdp_mergable, by = "iso3c") %>% 
  left_join(tfr_mergable, by = "iso3c")  
```

Taiwan data
```{r}
Pop_2019_Taiwan <- pop1dt %>%
  filter(year == 2019) %>%
  filter(country_code == 158) %>% 
  mutate(pop = pop * 1000) %>% 
  select(country_code, name, pop, year)
Pop_2019_Taiwan <- Pop_2019_Taiwan$pop
```
```{r, Taiwan}
# https://eng.stat.gov.tw/Point.aspx?sid=t.9&n=4208&sms=11713
# population_2021_Taiwan <- 23361084 # now using UN WPP data instead
# https://eng.stat.gov.tw/cp.aspx?n=2334
GDP_pp_Taiwan <- 32474
# https://population.un.org/dataportal/countryProfiles/types/1/topics/5/coreThemes/1/locations/158?classId=reg&palette=Blues&secpalette=alphabet2
# https://www.cia.gov/the-world-factbook/countries/taiwan/ # 75.2
Contracept_taiwan <- 74.7
# https://www.nature.com/articles/s41598-021-94540-7
Footprint_pp_Taiwan <- 6.46
Biocapacity_pp_Taiwan <- 1.3
# for Sustainable Pop calculation
Sustainable_pop_Taiwan <- floor(Biocapacity_pp_Taiwan / Footprint_pp_Taiwan * Pop_2019_Taiwan)
```

Taiwan insert data in countries table
```{r}
eo_2023[eo_2023$iso3c == "TWN", "Contraception"] = Contracept_taiwan
# eo_2023[eo_2023$iso3c == "TWN", "Year_contracept"] = # not supplied
eo_2023[eo_2023$iso3c == "TWN", "GDP"] = GDP_pp_Taiwan
eo_2023[eo_2023$iso3c == "TWN", "Year_gdp"] = 2022
eo_2023[eo_2023$iso3c == "TWN", "Population_gfn_2019"] = Pop_2019_Taiwan
eo_2023[eo_2023$iso3c == "TWN", "Footprint_pp_2019"] = Footprint_pp_Taiwan
eo_2023[eo_2023$iso3c == "TWN", "Biocapacity_pp_2019"] = Biocapacity_pp_Taiwan
eo_2023[eo_2023$iso3c == "TWN", "Sustainable_pop"] = Sustainable_pop_Taiwan
```

Archive and export eo_2023
```{r}
eo_2023_order <- eo_2023 %>% 
  rename("Population_2019" = "Population_gfn_2019") %>% 
  select("iso3c", "Country", "Photos", "Biocapacity_pp_2019", "Footprint_pp_2019", "Population_2019", "Sustainable_pop", "PopGrowRate_UN_2021", "Population_UN_2021", "Species_2022_2", "Contraception", "Year_contracept",   "GDP", "Year_gdp", "TFR", "Continent", "iso2c")
```

Sustainability Grade column
as of July 2023 still the same as in November
```{r}
grades <- eop_26April2022 %>% 
  select(iso3c, Rank_sustain_2020)

eo_2023_beforeGrade <- eo_2023_order # preserve

eo_2023 <- eo_2023_order %>% 
  left_join(grades, by = "iso3c") %>% 
  rename(Grade = Rank_sustain_2020)
```

Merge in World Bank population 2022 and popgrow rate 2022
```{r}
wbpop2022_mergable <- WB_pop_2022 %>% 
  rename(Population_WB_2022 = SP.POP.TOTL) %>% 
  select(iso3c, Population_WB_2022)
  
wbpopgrow2022_mergable <- popgrow_2022 %>% 
  rename(PopGrowRate_WB_2022 = SP.POP.GROW) %>% 
  select(iso3c, PopGrowRate_WB_2022)

# eo_before_WB_pop <- eo_2023
eo_2023 <- eo_before_WB_pop %>% 
  left_join(wbpop2022_mergable, by = "iso3c") %>%
  left_join(wbpopgrow2022_mergable, by = "iso3c")
```

Delete labels, 3 decimal places for WB popgrow, reorder columns
```{r}
attr(eo_2023$GDP, "label") <- NULL
attr(eo_2023$Contraception, "label") <- NULL
attr(eo_2023$Population_WB_2022, "label") <- NULL
attr(eo_2023$PopGrowRate_WB_2022, "label") <- NULL

eo_2023$PopGrowRate_WB_2022 <- round(eo_2023$PopGrowRate_WB_2022, 3)

eo_2023 <- eo_2023 %>% 
  select("iso3c", "Country", "Grade", "TFR", "Photos", "Sustainable_pop", "Species_2022_2", "Population_WB_2022", "PopGrowRate_WB_2022", "Population_UN_2021", "PopGrowRate_UN_2021", "Contraception", "Year_contracept", "GDP", "Year_gdp", "Continent", "Biocapacity_pp_2019", "Footprint_pp_2019", "Population_2019", "iso2c")
```

Taiwan copy UN population and growrate into World Bank columns 
```{r}
Taiwan_population_UN_2021 <- eo_2023[165, 10] 
eo_2023[eo_2023$iso3c == "TWN", "Population_WB_2022"] = Taiwan_population_UN_2021
Taiwan_popgrow_UN_2021 <- eo_2023[165, 11] 
eo_2023[eo_2023$iso3c == "TWN", "PopGrowRate_WB_2022"] = Taiwan_popgrow_UN_2021

# French Guiana[61] missing in WB, present in UN
eo_2023[61,8] <- eo_2023[61,10]
eo_2023[61,9] <- eo_2023[61,11]

# Bulgaria WB 2022 popgrow -6.2% replaced by UN 2021
eo_2023[25,9] <- eo_2023[25,11]
```

Round off numbers
```{r}
# dput(names(eo_2023)) # gets column names
# eo_before_rounding <- eo_2023
# eo_2023 <- eo_before_rounding %>% 
#   select()
```

Archive and export for web sheet
```{r}
saveRDS(eo_2023, file="data/eo_WB_2023.rds")
write_csv(eo_2023, "data/eo_WB_2023.csv")
```

